---
title: "Practical Machine Learning Class - Prediction Assignment Writeup"
author: "Hazm Talab"
date: "Wednesday, July 23, 2014"
output: html_document
---
## Abstract  
In this assignment we are trying to build a predictive model to predict human activity using different type of measurements data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants.

## Getting the Data  
The data for this project come from this source: <http://groupware.les.inf.puc-rio.br/har>.  
The training data for this project are available here: 
<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv>  
The test data are available here: 
<https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv>  

## Loading the Data  

```{r LoadingData}
pml.training <- read.csv("pml-training.csv")
pml.testing <- read.csv("pml-testing.csv")
```

## Exploring the data

```{r ExploringData}
dim(pml.training)
dim(pml.testing)
str(pml.training)
```

We notice that there are many numerical variables treated as factor variables 
because they have "#DIV/0!" error so we reload the data and pass this string to 
na.strings parameter to consider this error a missing value.

```{r ReloadingData}
pml.training <- read.csv("pml-training.csv", na.strings = c("NA","#DIV/0!"))
pml.testing <- read.csv("pml-testing.csv", na.strings = c("NA","#DIV/0!"))
str(pml.training)
```

There still many missing values. So, in order to clean the data we examine the features that 
has missing values and drop them out of the data set. In other words, we only keep the features 
which do not have any missing values. We also drop out the first seven features, which are not much relevant.

```{r RemoveNAFeatures}
data <- pml.training[colnames(pml.training[colSums(is.na(pml.training)) == 0])][-(1:7)]
dim(data)
```

We notice that we dropped out more than 100 features. That leave us with `r dim(data)[2]-1` predictors, and of course that will help us build our model more efficiently.  

## Creating training and testing datasets
Let us now split the dataset into two datasets using **caret** package.  

```{r CreatingTrainTest}
library(caret)
set.seed(1234)
inTrain <- createDataPartition(data$classe, p=0.7, list=FALSE)
training <- data[inTrain,]
testing <- data[-inTrain,]
```

## Creating a model Fit
We use **randomForest** package with **caret** package to build our prediction model using random forest algorithm with cross validation. We use only the raining dataset we created.  
We use **doParallel** package to use the CPU cores for the computations.   

```{r CreatingModel, cache=TRUE}
library(randomForest)
library(doParallel)
cl <- makeCluster(detectCores())
registerDoParallel(cl)
modFit <- train(classe ~ ., data=training, method="rf", trControl = trainControl(method = 'cv'))
stopCluster(cl)
````

## Evaluating the model
We evaluate the model on the training dataset first.  
```{r EvaluatingModelTRAIN}
TrainingPredictions <- predict(modFit, newdata=training)
confusionMatrix(TrainingPredictions, training$classe)
```
We notice from the result of confusionMatrix that fit the data perfectly with accuracy of 
`r confusionMatrix(TrainingPredictions, training$classe)$overall[1]`.  

Then we evaluate the model on the testing dataset we create.  
```{r EvaluatingModelTEST}
TestingPredictions <- predict(modFit, newdata=testing)
confusionMatrix(TestingPredictions, testing$classe)
```
The results of confusionMatrix shows that the model is efficient. It fit the testing dataset very well. The accuracy is `r confusionMatrix(TestingPredictions, testing$classe)$overall[1]`.   

## The estimated out of sample error
As we expected the estimated out of sample error is very low using random forest algorithm with cross validation.
```{r FinalModel}
modFit$finalModel
```


## Predicting the test cases provided by Coursera
Finally we predict the outcome on **pml.testing** dataset, the test dataset provided by Coursera that we loaded early, we test the 20 cases and check using Coursera dashboard. We predict 20 out 20 correctly.  

We use the following code to predict the outcome.  
```{r PredictTEST}
answers = predict(modFit, newdata=pml.testing)
answers
```

We use the following function and code to generate the files we submitted to Coursera.
```{r WriteAnswers, eval=FALSE}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
setwd("Answers/")
pml_write_files(answers)
```

